Modelling the real-time state of a public transport road network

### 2017-05-02 realtime-filter/realtime-filter.Rnw
fix forumulae
things
kf results + pictures

\title{Modelling the Real Time State of a Public Transport Road Network}
\author{Tom~Elliott}

@@

\begin{abstract}
  Model all vehicles in the public transport system in real time.
  Particle filter model for each vehicle (bus),
  using observations of position (GPS coordinates) to infer distance into trip
  and velocity.
  Then use particle filter estimates of vehicle speed to estimate bus speed along
  road segments, independently of route.

@@

%\subsubsection[L1b]{$L_1'$: at stop}
\item
If the particle is considered to be at a stop, then we apply the following likelihood.
Define a parameter $\varepsilon_S$ as a boundary around the stop;
if observation is within that boundary likelihood is 1, else 0.
Let $B_s$ be a circle with radius $\varepsilon_S$ and centered at stop $j$, then

@@

%\subsubsection[L1c]{$L_1''$: at intersection}
\item
If the particle is considered to be at an intersection, we apply the following likelihood.
Similar to $L_1'$, however need to check observation is heading into the intersection
(and not out of it) by using two boundaries around the intersection and a point $\varepsilon_I$ away.
Let $B_I^1$ and $B_I^2$ be two circles with radii $\varepsilon_I$,
the first centered on the intersection and the second centered $\varepsilon_I$ meters before it, then

@@

%\subsubsection[L2]{$L_2$: not at a stop}

@@

\item
  If $s_k^{(i)} \neq j$ (particle's stop index isn't the same as the stop time update's stop sequence),
  then we check particle is plausible given the data:
  \begin{equation}
    \label{eq:l2a}
    L_2 (\bT_k | \bX_k^{(i)}) =
    \mathbbm{1}_{s_k^{(i)} > j\ \cap\ A_{s_k}^{(i)} > T_j^a\ \cap\ A_{s_k}^{(i)} > T_j^d},
  \end{equation}
  (defining $x > y$ as TRUE if either $x$ or $y$ are NULL to simplify notation); else

@@

We run the Kalman filter every time we run the particle filter,
since it has a low cost, allowing the most recent data to be incorporated.
We have $N$ weighted particle estimates of vehicle state for each bus,
with which we can compute the mean distance into trip and velocity for bus $j$,
\begin{equation}
  \label{eq:weighted_mean}
  \bar d_j = \sum_{i=1}^N w_j^{(i)} d_j^{(i)} \quad\text{and}\quad
  \bar v_j = \sum_{i=1}^N w_j^{(i)} v_j^{(i)},
\end{equation}
as well as the weighted sample variance \cite{cn},
\begin{equation}
  \label{eq:weighted_variance}
  \bar S_j^2 = \frac{\sum_{i=1}^N w_j^{(i)} (v_j^{(i)} - \bar v_j)^2}{1 - \sum_{i=1}^N (w_j^{(i)})^2},
\end{equation}
since $\sum_{i=1}^N w_j^{(i)} = 1\quad \forall j$.

@@

@@

>>>>

\title{Modelling the real-time state of a public transport road network}
\author{Tom~Elliott}

@@

\begin{abstract}
  Model all vehicles in the public transport system in real-time.
  Particle filter model for each vehicle (bus),
  using observations of position (GPS coordinates) to infer distance into trip
  and velocity.
  Then use particle filter estimates of vehicle speed to estimate bus speed along
  road segments, independently of route.

@@

%\subsubsection[L1b]{$L_1'$: at stop}
\item
If the particle is considered to be at a stop, then we apply the following likelihood.
Define a parameter $\varepsilon_S$ as a boundary around the stop;
if observation is within that boundary likelihood is 1, else 0 (figure~\ref{fig:likelihood_circles}a).
Let $B_s$ be a circle with radius $\varepsilon_S$ and centered at stop $j$, then

@@

%\subsubsection[L1c]{$L_1''$: at intersection}
\item
If the particle is considered to be at an intersection, we apply the following likelihood.
Similar to $L_1'$, however need to check observation is heading into the intersection
(and not out of it) by using two boundaries around the intersection and a point $\varepsilon_I$ away
(figure~\ref{fig:likelihood_circles}b).
Let $B_I^1$ and $B_I^2$ be two circles with radii $\varepsilon_I$,
the first centered on the intersection and the second centered $\varepsilon_I$ meters before it, then

@@

<<likelihood_circles,echo=FALSE,out.width='0.5\\linewidth',fig.cap="Likelihood functions for particle's at stops and intersections. The grey region represents the area where particle's are accepted.",fig.subcap=c("Stop", "Intersection")>>=
par(mar = rep(2.1, 4))
plot(NA, bty = "n", type = "n", xaxt = "n", yaxt = "n", xlim = 0:1, ylim = 0:1, asp = 1)
lines(c(0, 1), c(0.5, 0.5))
th <- seq(0, 360, length = 100) * pi / 180
polygon(cos(th) / 4 + 0.5, sin(th) / 4 + 0.5, col = "#cccccc")
lines(c(0, 1), c(0.5, 0.5), lty = "9999")
points(0.5, 0.5, pch = 19, cex = 4)

plot(NA, bty = "n", type = "n", xaxt = "n", yaxt = "n", xlim = 0:1, ylim = 0:1, asp = 1)
lines(c(0, 1), c(0.5, 0.5))
th <- seq(0, 360, length = 100) * pi / 180
th2 <- seq(-180, 180, length = 100) * pi / 180
lines(c1x <- cos(th) / 4 + 0.7, c1y <- sin(th) / 4 + 0.5, lty = "99")
lines(c2x <- cos(th2) / 4 + 0.7 - 0.25, c2y <- sin(th2) / 4 + 0.5, lty = "99")
wc1 <- c1x <= 0.7 - 0.25/2
wc2 <- c2x >= 0.7 - 0.25/2
polygon(c(c1x[wc1], c2x[wc2]), c(c1y[wc1], c2y[wc2]), col = "#cccccc")
lines(c(0, 1), c(0.5, 0.5), lty = "9999")
points(0.7, 0.5, pch = 19, cex = 4)
points(0.7 - 0.25, 0.5, pch = 15, cex = 3)
@

@@

\item
  If $s_k^{(i)} \neq j$ (particle's stop index isn't the same as the stop time update's stop sequence),
  then we check particle is plausible given the data:
  \begin{equation}
    \label{eq:l2a}
    L_2 (\bT_k | \bX_k^{(i)}) =
    \mathbbm{1}_{s_k^{(i)} > j\ \cap\ A_{s_k}^{(i)} > T_j^a\ \cap\ A_{s_k}^{(i)} > T_j^d\
      \cap\ T^a_j > t_k\ \cap\ T^d_j > t_k}
  \end{equation}
  (defining $x > y$ as TRUE if either $x$ or $y$ are NULL to simplify notation); else

@@

We run the Kalman filter every time we run the particle filter,
since it has a low cost, allowing the most recent data to be incorporated.
We have $N$ weighted particle estimates of vehicle state for each bus,
with which we can compute the mean distance into trip and velocity for bus $j$,
\begin{equation}
  \label{eq:weighted_mean}
  \bar d_j = \sum_{i=1}^N w_j^{(i)} d_j^{(i)} \quad\text{and}\quad
  \bar v_j = \sum_{i=1}^N w_j^{(i)} v_j^{(i)},
\end{equation}
as well as the weighted sample variance \cite{cn},
\begin{equation}
  \label{eq:weighted_variance}
  \bar S_j^2 = \sum_{i=1}^N w_j^{(i)} (v_j^{(i)} - \bar v_j)^2,
\end{equation}
since $\sum_{i=1}^N w_j^{(i)} = 1\quad \forall j$.

@@

The transport network state is composed of $L$ segments,
defined by the mean speed in each segment, 
\mbox{$\bnu = [\nu_1, \nu_2, \cdots, \nu_L]^T$},
and the variance of speed within each segment, $\bxi$, 
a diagonal matrix where the $\xi_{\ell\ell}$ is the variance of the segment $\ell$.
The Kalman filter predict step is
\begin{align}
  \bnu_{k|k-1} &= \bnu_{k-1|k-1} \\
  \bxi_{k|k-1} &= \bxi_{k-1|k-1} + \bQ_k
\end{align}
where $\bQ_k \sim \mathcal{N}(0, \Delta_k\sigma_{\nu}^2I_L)$ and $\sigma_{\nu}^2$ is the system noise parameter
and $I_n$ is an $n\times n$ identity matrix.

The update equations simplify to
\begin{align}
  \bK_k &= \frac{\bxi_{k|k-1}}{\bxi_{k|k-1} + \hat R^2} \\
  \nu_{k|k} &= \bnu_{k|k-1} + \bK_k(\bar\bz_k - \bnu_{k|k-1}) \\
  \bxi_{k|k} &= (I_L - \bK_k) \bxi_{k|k-1}
\end{align}
which make use of the segment means and variances computed in (\ref{eq:kf_data_estimates}).

@@

- model implemented in C++

### 2017-05-02 realtime-filter/realtime-filter.tex
fix forumulae
things
kf results + pictures

\title{Modelling the Real Time State of a Public Transport Road Network}
\author{Tom~Elliott}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}

@@

%\subsubsection[L2]{$L_2$: not at a stop}

>>>>

\title{Modelling the real-time state of a public transport road network}
\author{Tom~Elliott}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
\subfloat[Stop\label{fig:likelihood_circles1}]{\includegraphics[width=0.5\linewidth]{figure/likelihood_circles-1} }
\subfloat[Intersection\label{fig:likelihood_circles2}]{\includegraphics[width=0.5\linewidth]{figure/likelihood_circles-2} }\caption[Likelihood functions for particle's at stops and intersections]{Likelihood functions for particle's at stops and intersections. The grey region represents the area where particle's are accepted.}\label{fig:likelihood_circles}
\end{figure}
